---
title: "Analysis Process"
author: "Anthony Contoleon"
date: "6 March 2017"
output:
  github_document: default
  html_document: default
---

# Analysing Australian Preference Distribution

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
## 2016 preference flow to major parties.
## Data sources
# http://www.aec.gov.au/Elections/federal_elections/Stats_CDRom.htm
# http://results.aec.gov.au/20499/Website/HouseDownloadsMenu-20499-Csv.htm

## Analysis and display
library(ggplot2)
library(reshape2)
library(plyr)

library(MASS)
library(ca)
library(FactoMineR)
library(factoextra)

library(smacof)

library(ggbiplot)
library(scales)
library(grid)

library(caret)

library(dplyr)
library(gplots)
library(colorspace)
library(dendextend)
library(cluster) 
library(fpc)

## PCA analysis chart function
## https://www.r-bloggers.com/computing-and-visualizing-pca-in-r/
## http://rstudio-pubs-static.s3.amazonaws.com/27823_dbc155ba66444eae9eb0a6bacb36824f.html
pcaCharts <- function(x) {
  x.var <- x$sdev ^ 2
  x.pvar <- x.var/sum(x.var)
  print("proportions of variance:")
  print(x.pvar)
  
  par(mfrow=c(2,2))
  plot(x.pvar,xlab="Principal component", ylab="Proportion of variance explained", ylim=c(0,1), type='b')
  plot(cumsum(x.pvar),xlab="Principal component", ylab="Cumulative Proportion of variance explained", ylim=c(0,1), type='b')
  screeplot(x)
  screeplot(x,type="l")
  par(mfrow=c(1,1))
}

## Function for exporting tables to CSVs. 
##Directory path other than work directory and extension needs to be defined 'x', 'y' is the table to be exported.
file_output <- function (x, y){
  path <- paste(work_dir, x, sep ="")
  write.table(y, file = path, sep = ",", row.names = FALSE)
}

## The summary statistcs by group.
summaryStatistics <- function(x) {
  c(min = min(x), max = max(x), 
    mean = mean(x), median = median(x), 
    std = sd(x))
}

## Setting date
date <- format(Sys.Date(), "%Y%m%d")

## Get working directory path
setwd("C:/data/other-projects/data/firstPreferences/")
work_dir <- getwd()

## Import the pre-processed file produced from dataImport.R
reImport.df <- read.csv("20170129-electionPreferences.csv", stringsAsFactors = FALSE)

## Select the appropriate columns.
preferencesWip.df <- reImport.df
preferencesWip.df$FromCandidatePartyAb <- ifelse(preferencesWip.df$FromCandidatePartyAb == "", paste("FP", preferencesWip.df$ToCandidatePartyAb, sep = " "), preferencesWip.df$FromCandidatePartyAb)
preferencesWip.df <- na.omit(preferencesWip.df)
preferencesWip.df[sapply(preferencesWip.df, is.character)] <- lapply(preferencesWip.df[sapply(preferencesWip.df, is.character)], as.factor)

## Removing first preference columns
preferencesWipExplore.df <- preferencesWip.df[-grep("FP ", preferencesWip.df$FromCandidatePartyAb),]

## Calculating percentages.
## State level
preferencesWipExploreNat.df <- ddply(preferencesWipExplore.df, .(StateAb, ToCandidatePartyAb, electionYear), transform, percentVotesStateTo = TransferCount/sum(TransferCount)*100)

preferencesWipExploreNat.df <- ddply(preferencesWipExplore.df, .(StateAb, FromCandidatePartyAb, electionYear), transform, percentVotesStateFrom = TransferCount/sum(TransferCount)*100)

## National level
preferencesWipExploreState.df <- ddply(preferencesWipExplore.df, .(FromCandidatePartyAb, electionYear), transform, percentVotesOverallFrom = TransferCount/sum(TransferCount)*100)

preferencesWipExploreState.df <- ddply(preferencesWipExplore.df, .(ToCandidatePartyAb, electionYear), transform, percentVotesOverallTo = TransferCount/sum(TransferCount)*100)

## Electorate level
preferencesWipExplore.df <- ddply(preferencesWipExplore.df, .(FromCandidatePartyAb, DivisionNm, electionYear), transform, percentVotesDisctFrom = TransferCount/sum(TransferCount)*100)

preferencesWipExplore.df <- ddply(preferencesWipExplore.df, .(ToCandidatePartyAb, DivisionNm, electionYear), transform, percentVotesDisctTo = TransferCount/sum(TransferCount)*100)

```


## Exploration of the data set

Preference flows at Queensland and National level for the Greens and One Nation.

### QLD Preference Flows

```{r explorationState, echo=FALSE, message=FALSE, warning=FALSE}
## Producing plots.
ggplot(preferencesWipExplore.df[which(preferencesWipExplore.df$StateAb=='QLD' & preferencesWipExplore.df$FromCandidatePartyAb=='ON'),], aes(x = DivisionNm, y = percentVotesDisctFrom, fill = ToCandidatePartyAb)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + facet_grid(electionYear ~ .) + facet_grid(electionYear ~ .) + scale_fill_manual(values=c("#d73027", "#4575b4", "#74add1", "#fee090")) + ggtitle("QLD One Nation Party Two Candidate Preference Flows")

ggplot(preferencesWipExplore.df[which(preferencesWipExplore.df$StateAb=='QLD' & preferencesWipExplore.df$FromCandidatePartyAb=='GRN'),], aes(x = DivisionNm, y = percentVotesDisctFrom, fill = ToCandidatePartyAb)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + facet_grid(electionYear ~ .) + facet_grid(electionYear ~ .) + scale_fill_manual(values=c("#b2182b","#6a51a3", "#d73027", "#4575b4", "#4575b4","#878787","#ffed00")) + ggtitle("QLD Green Party Two Candidate Preference Flows")
```

### National Preference Flows

```{r explorationnational, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(preferencesWipExplore.df[which(preferencesWipExplore.df$FromCandidatePartyAb=='ON'),], aes(x = DivisionNm, y = percentVotesDisctFrom, fill = ToCandidatePartyAb)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + facet_grid(electionYear ~ .) + facet_grid(electionYear ~ .) + scale_fill_manual(values=c("#b2182b","#999999", "#aaaaaa", "#6a51a3", "#4575b4", "#4575b4", "#4575b4","#1a9850","#ffed00")) + ggtitle("National One Nation Party Two Candidate Preference Flows")

ggplot(preferencesWipExplore.df[which(preferencesWipExplore.df$FromCandidatePartyAb=='GRN'),], aes(x = DivisionNm, y = percentVotesDisctFrom, fill = ToCandidatePartyAb)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0)) + facet_grid(electionYear ~ .) + facet_grid(electionYear ~ .) + scale_fill_manual(values=c("#b2182b","#999999", "#aaaaaa", "#6a51a3", "#d73027", "#4575b4","#4575b4","#4575b4", "#1a9850","#878787","#ffed00","#FF5800")) + ggtitle("National Green Party Two Candidate Preference Flows")
```


## Average ALP share of preferences.

### Percentage of votes to ALP from GRN
```{r tablesALP1, echo=FALSE, message=FALSE, warning=FALSE}
wipGRN.df <- aggregate(percentVotesDisctFrom ~ DivisionNm + electionYear, data = preferencesWipExplore.df[which(preferencesWipExplore.df$FromCandidatePartyAb=='GRN' & preferencesWipExplore.df$ToCandidatePartyAb=='ALP'),], sum)
tapply(wipGRN.df$percentVotesDisctFrom, wipGRN.df$electionYear, summaryStatistics)

```

### Percentage of votes to ALP from ON
```{r tablesALP2, echo=FALSE, message=FALSE, warning=FALSE}
## Average APL share of preferences.
wipONA.df <- aggregate(percentVotesDisctFrom ~ DivisionNm + electionYear, data = preferencesWipExplore.df[which(preferencesWipExplore.df$FromCandidatePartyAb=='ON' & preferencesWipExplore.df$ToCandidatePartyAb=='ALP'),], sum)
tapply(wipONA.df$percentVotesDisctFrom, wipONA.df$electionYear, summaryStatistics)
```

## Average LNP/LP share of preferences.

### Percentage of votes to LNP/LP from ON
```{r tablesLP, echo=FALSE, message=FALSE, warning=FALSE}
wipONL.df <- aggregate(percentVotesDisctFrom ~ DivisionNm + electionYear, data = preferencesWipExplore.df[which(preferencesWipExplore.df$FromCandidatePartyAb=='ON' & grepl("^LNP|^LNQ|^LP$", preferencesWipExplore.df$ToCandidatePartyAb)),], sum)
tapply(wipONL.df$percentVotesDisctFrom, wipONL.df$electionYear, summaryStatistics)
```

## Processing

Prior to further analysis the data was processed to re-label LNP, LP and LNQ to LP. The data was subsetted to only include the 2016 federal election and parties with fewer than 10,000 votes were removed from the data set. The data set was transformed to be appropriate for use with the packages used in the following techniques.

```{r processPCACA, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
preferencesWipExplore.df[which(grepl("^LNP|^LNQ|^LP$", preferencesWipExplore.df$FromCandidatePartyAb)),]

## Electorate level with first preferences
preferencesWipMDS.df <- ddply(preferencesWip.df, .(ToCandidatePartyAb, DivisionNm, electionYear), transform, percentVotesDisctTo = TransferCount/sum(TransferCount)*100)
preferencesWipMDS.df <- ddply(preferencesWipMDS.df, .(FromCandidatePartyAb, DivisionNm, electionYear), transform, percentVotesDisctFrom = TransferCount/sum(TransferCount)*100)
preferencesWipExplore.mds <- preferencesWipMDS.df[c(1,3,7,8,13,14,18,21,22,23)]

preferencesWipExploreAg.mds <- aggregate(cbind(percentVotesDisctFrom, percentVotesDisctTo, TransferCount) ~ ToCandidatePartyAb + FromCandidatePartyAb + DivisionNm + electionYear + StateAb + FromCandidatePartyNm + ToCandidatePartyNm, data = preferencesWipExplore.mds, sum)

## Loading and preparing the data
preferences2016.df <- preferencesWipExploreAg.mds[which(preferencesWipExploreAg.mds$electionYear==2016),]
preferences2016.mds <- preferences2016.df[,c(1,2,8)]
check2016.df <- preferences2016.df
check2016.mds <- aggregate(TransferCount ~ FromCandidatePartyAb, data = preferences2016.df, sum)

## Minor parties to exclude
exclude <- check2016.mds[which(check2016.mds$TransferCount < 10000),]$FromCandidatePartyAb
check2016.df <- check2016.df[!(check2016.df$FromCandidatePartyAb %in% exclude),]

## Creating the frame for use in the analysis
preferences2016.ca <- check2016.df[!grepl("^FP ", check2016.df$FromCandidatePartyAb), ]
preferences2016.ca[preferences2016.ca$ToCandidatePartyAb == "LNP",] <- "LP"
preferences2016.ca[preferences2016.ca$FromCandidatePartyAb == "LNP",] <- "LP"
preferences2016.ca$TransferCount <- as.numeric(preferences2016.ca$TransferCount)
preferences2016.ca$percentVotesDisctTo <- as.numeric(preferences2016.ca$percentVotesDisctTo)
preferences2016.ca$percentVotesDisctFrom <- as.numeric(preferences2016.ca$percentVotesDisctFrom)
preferences2016.ca[is.na(preferences2016.ca)] <- 0

preferences2016ag.ca <- aggregate(TransferCount ~ ToCandidatePartyAb + FromCandidatePartyAb, data = preferences2016.ca, mean)

## Create the data frame for use with the correspondence analysis
correspondencePref.df <- dcast(preferences2016ag.ca, FromCandidatePartyAb ~ ToCandidatePartyAb, sum, value.var = "TransferCount")
rownames(correspondencePref.df) <- paste0("From_", correspondencePref.df$FromCandidatePartyAb)
colnames(correspondencePref.df) <- paste0("To_", colnames(correspondencePref.df))
rowNum <- dim(correspondencePref.df)[1]
colNum <- dim(correspondencePref.df)[2]
correspondencePref.df <- correspondencePref.df[c(1:rowNum),c(2:colNum)]

## PCA Analysis of the data.
# log transform 
correspondencePrefLog.df <- log(correspondencePref.df[,c(1,3,6,7)])
parties.df <- rownames(correspondencePrefLog.df)

## Replacing infinite values with zero
correspondencePrefLog.df <- do.call(data.frame,lapply(correspondencePrefLog.df, function(x) replace(x, is.infinite(x),0)))

# apply PCA - scale. = TRUE is highly 
# advisable, but default is FALSE. 
correspondencePrefLog.pca <- prcomp(correspondencePrefLog.df,
                                    retx = TRUE,
                                    center = TRUE,
                                    scale. = TRUE) 
```

## PCA Process

Performing a Principle Componant Analysis (PCA) on the supplied and transformed data.

```{r PCA, echo=FALSE, message=FALSE, warning=FALSE}
# plot method
plot(correspondencePrefLog.pca, type = "l")

# summary method
summary(correspondencePrefLog.pca)

# Predict PCs
predict(correspondencePrefLog.pca, 
        newdata = tail(correspondencePrefLog.df, 10))
```

### PCA Plots

Additional exploration was performed on the data. The following are the stock plots.

```{r PCAdetail, echo=FALSE, message=FALSE, warning=FALSE}
## Stock plots
pcaCharts(correspondencePrefLog.pca)
biplot(correspondencePrefLog.pca, scale = 0, cex = .7)
```

### Additional Plots

Plot was produced using the 'ggbiplot' package, currently not available on CRAN.

The placement of the ALP and LP within this graph is more a product of the rarity of the situations where preferences would be recorded to flow from one of the major parties. With the dominance of the major parties within the House of Representatives, it is not common that they both would not be present as the recipiant of the preferences. The most common occurances where this occures are addressed with the inclusion of both GRN and NP as variables within the model, though this won't address the small number of observations.

```{r PCAdetailAdditional, echo=FALSE, message=FALSE, warning=FALSE}
#library(devtools)
#install_github("ggbiplot", "vqv")

g <- ggbiplot(correspondencePrefLog.pca, obs.scale = 1, var.scale = 1, labels = row.names(correspondencePref.df),
              ellipse = TRUE, 
              circle = TRUE)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)

trans = preProcess(correspondencePref.df, 
                   method=c("BoxCox", "center", 
                            "scale", "pca"))
PC = predict(trans, correspondencePref.df)
```

### Retained Principle Componants

```{r PCAretanied, echo=FALSE, message=FALSE, warning=FALSE}
# Retained PCs
head(PC, 3)
```

## CA Process

Performing a Correspondence Analysis (CA) on the supplied and transformed data.

```{r CA, echo=FALSE, message=FALSE, warning=FALSE}
## Column cut down.
suppressRow <- grep("From_FP.*", rownames(correspondencePref.df))
suppressCol <- c(2,4,5,8,9)
#correspondencePref.df <- correspondencePref.df[c(1,3,5,6,7,8,9,10)]

## setting values for fit
chisq <- chisq.test(correspondencePref.df)
n <- sum(correspondencePref.df)
chisq

## Total inertia value
chisq$statistic/n

# Removing columns with all zero.
correspondencePref.df <- correspondencePref.df[, !apply(is.na(correspondencePref.df) | correspondencePref.df == 0, 2, all)]
correspondencePref.df <- correspondencePref.df[ rowSums(correspondencePref.df[c(1,5)])!=0, ] 

## Principal intertias table/contingency table
# Performing the analysis using two different packages to ensure that all the data required as per SPSS examples is generated.
#survey.ca <- ca(correspondencePref.df, nd=2, suprow = suppress)
#survey.cac <- CA(correspondencePref.df, ncp = 2, graph = TRUE, row.sup = suppress)
survey.ca <- ca(correspondencePref.df, nd = 3, supcol = suppressCol)
survey.cac <- CA(correspondencePref.df, ncp = 3, graph = TRUE, col.sup = suppressCol)
survey.ca

## Contribution figures for effect per row/col
# Displayed as percent
survey.cac$row$contrib
survey.cac$col$contrib
```

## Plot combinations of dimensions.

Two plots covering the three different dimensions used.

```{r CAplot, echo=FALSE, message=FALSE, warning=FALSE}
fviz_ca_biplot(survey.ca, axes = c(1,2), geom="text", arrow = c(FALSE, TRUE))
fviz_ca_biplot(survey.ca, axes = c(1,3), geom="text", arrow = c(FALSE, TRUE))

```

## Clustering

K means clustering was performed with three clusters as per the results seen from PCA and the Hierachial cluster model with four.

```{r clustering, echo=FALSE, message=FALSE, warning=FALSE}
preference2016HCag.df <- aggregate(percentVotesDisctFrom ~ ToCandidatePartyAb + FromCandidatePartyAb, data = preferences2016.ca, mean)

## Create the data frame for use with the correspondence analysis
preference2016HCdcast.df <- dcast(preference2016HCag.df, FromCandidatePartyAb ~ ToCandidatePartyAb, sum, value.var = "percentVotesDisctFrom")
rownames(preference2016HCdcast.df) <- paste0("From_", preference2016HCdcast.df$FromCandidatePartyAb)
colnames(preference2016HCdcast.df) <- paste0("To_", colnames(preference2016HCdcast.df))

preference2016KMdcast.df <- dcast(preference2016HCag.df, ToCandidatePartyAb ~ FromCandidatePartyAb, sum, value.var = "percentVotesDisctFrom")
rownames(preference2016KMdcast.df) <- paste0("TO_", preference2016KMdcast.df$ToCandidatePartyAb)
colnames(preference2016KMdcast.df) <- paste0("FROM_", colnames(preference2016KMdcast.df))

## Subsetting the data set. AS the data is in percentages, scaling it not needed.
x <- preference2016HCdcast.df[c(2,4,7,8)]
```

### K Means

Three groups were used for k means clustering based on the initial PCA. Each group can be described based on the parties with whom the preferences flowed to from the member parties. Group 1 is prediinantly ALP, with LP being the next most important contributing factor. Group 2 has both LP and NP conributing almost the same, and represents what can be considered the right side of Australian politics. Group 3 is predimantly defined by GRN with ALP and LP next in that order. This group appears to mostly catch those who are voting for center to left minor parties.

Placement of the ALP, LP and NP in these plots themselves are more an artifact of the way that the preference voting system works and how the two party preferred statistics are produced.

```{r kmeans, echo=FALSE, message=FALSE, warning=FALSE}
kmx.df <- x

# K-Means Cluster Analysis
fit <- kmeans(kmx.df, 3)
# get cluster means 
aggregate(kmx.df,by=list(fit$cluster),FUN=mean)
# append cluster assignment
kmx.df <- data.frame(kmx.df, fit$cluster)

# vary parameters for most readable graph
clusplot(kmx.df, fit$cluster, color=TRUE, shade=TRUE, 
         labels=2, lines=0)

```

### Hierachial

```{r hierachial, echo=FALSE, message=FALSE, warning=FALSE}
hcx.df <- x
## Hierachial Clustering
hc.complete <- hclust(dist(hcx.df, method = "euclidean"), method = "complete")
hc.average <- hclust(dist(hcx.df, method = "euclidean"), method = "average")
hc.single <- hclust(dist(hcx.df, method = "euclidean"), method = "single")
hc.ward <- hclust(dist(hcx.df, method = "euclidean"), method = "ward.D")
```

Ploting out the different hieracial models. The models are constructed with the assumption of three clusters. Each red rectangle defines the group. 

```{r hierachialPlot, echo=FALSE, message=FALSE, warning=FALSE}
plot(hc.complete, main = "Complete linkage", xlab = "", sub = "", cex = .9)
rect.hclust(hc.complete, k = 3, border="red")

plot(hc.average , main = "Average linkage", xlab = "", sub = "", cex = .9)
rect.hclust(hc.average, k = 3, border="red")

plot(hc.single , main = "Single linkage", xlab = "", sub = "", cex = .9)
rect.hclust(hc.single, k = 3, border="red")

plot(hc.ward , main = "Ward linkage", xlab = "", sub = "", cex = .9)
rect.hclust(hc.ward, k = 3, border="red")
```

Count of rows per group per model.

```{r hierachialTable, echo=FALSE, message=FALSE, warning=FALSE}
cluster.complete <- cutree(hc.complete, 3)
table(cluster.complete)
cluster.average <- cutree(hc.average, 3)
table(cluster.average)
cluster.single <- cutree(hc.single, 3)
table(cluster.single)
cluster.ward <- cutree(hc.ward, 3)
table(cluster.ward)

export.df <- preference2016HCdcast.df
export.df$tagK <- as.factor(kmx.df$fit.cluster)
export.df$tag <- as.factor(cutree(hc.complete, 4))
```

### Summary of variables

```{r hierachialDend, echo=FALSE, message=FALSE, warning=FALSE}
lapply(hcx.df, function(x) {
  if (is.numeric(x)) return(summary(x))
  if (is.factor(x)) return(table(x))
})

## https://cran.r-project.org/web/packages/dendextend/vignettes/Cluster_Analysis.html
dend <- as.dendrogram(hc.complete)
dend <- rotate(dend, 1:23)
dend <- color_branches(dend, k=6)
dend <- hang.dendrogram(dend,hang_height=0.1)

plot(dend, 
     main = "Clustered Claiments", 
     horiz =  TRUE,  nodePar = list(cex = .7))

## Preparing the data to be converted to a matrix.
# Define the function for the conversion
#factorNumeric <- function(x){as.numeric(levels(x))[x]}
factorNumeric <- function(x){as.numeric(factor(x))}
# Perform the conversion
facCol <- sapply(hcx.df, is.factor)
hcx.df[,facCol] <- lapply(hcx.df[,facCol], factorNumeric)

# Numeric scaling of the factors
hcx.df[facCol] <- lapply(hcx.df[facCol], scale)

labels_colors(dend) <-
  rainbow_hcl(6)[sort_levels_values(
    as.numeric(export.df$tag)[order.dendrogram(dend)]
  )]
```


### Dendrogram and Heatmap

Ploting the relative values of the matrix the previous cluster analyses were based on as a heat map. The heatmap below illustrates the relationships or lack of between recipiants of transfered preferences and those whom the preferences came from.

```{r hierachialHeatmap, echo=FALSE, message=FALSE, warning=FALSE}

# Heatmap graph
some_col_func <- function(n) rev(colorspace::heat_hcl(n, c = c(200, 60), l = c(60, 200), power = c(2, .2)))
gplots::heatmap.2(as.matrix(hcx.df), 
                  main = "Heatmap for the data set",
                  srtCol = 20,
                  dendrogram = "row",
                  Rowv = dend,
                  labRow = rownames(hcx.df),
                  Colv = "NA", # this to make sure the columns are not ordered
                  trace="none",          
                  #margins =c(5,0.1),      
                  key.xlab = "Scaled Value",
                  denscol = "grey",
                  density.info = "density",
                  #RowSideColors = rev(labels_colors(dend)),    
                  col = some_col_func
)


```

### Comparison of Groups

Comparing the two clustering methods by marking each group on a PCA plot. Both the From_ALP and From_LP points have proven to be problematic in this analysis. In part this was due to the very low number of times where preferences were allocated from a major party to a minor one.

### K Means

```{r clusterKM, echo=FALSE, message=FALSE, warning=FALSE}

g <- ggbiplot(correspondencePrefLog.pca, obs.scale = 1, var.scale = 1, labels = row.names(correspondencePref.df),
              groups = export.df$tag,
              ellipse = TRUE, 
              circle = TRUE)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)
```

### Hierachial Clustering

```{r clusterHC, echo=FALSE, message=FALSE, warning=FALSE}
g <- ggbiplot(correspondencePrefLog.pca, obs.scale = 1, var.scale = 1, labels = row.names(correspondencePref.df),
              groups = export.df$tagK,
              ellipse = TRUE, 
              circle = TRUE)
g <- g + scale_color_discrete(name = '')
g <- g + theme(legend.direction = 'horizontal', 
               legend.position = 'top')
print(g)

```